---
title: "Lab 3.1:Make a Plot"
subtitle: "Gabriela Chajon"
format: 
  revealjs:
    css: custom-style.css
editor: source
---

## Looking at the dataset
```{r}
library(datasetsICR)
```

```{r}
remotes::install_github("kjhealy/gssr")
```


```{r}
data("customers", package = "datasetsICR")
library(ggplot2)
head(customers, 14)
```

## First 10 customers
```{r}
head(customers, 10)
```
## Last 10 customers
```{r}
tail(customers, 10 )
```

## Summary statistics 
### Fresh statistics
```{r}
summary(customers$Fresh)
```
### Grocery statistics
```{r}
summary(customers$Grocery)
```

## Create a table for categorical variable
```{r}
customers$Region <- factor(customers$Region)
table(customers$Region)
```
## Create chart
```{r}
p <- ggplot(data = customers,
            mapping = aes(x = Fresh,
                          y = Grocery,
                          color = Region))
p + geom_point() + 
     scale_x_log10 (labels = scales::dollar) +
  scale_y_log10 (labels = scales::dollar) +
   geom_smooth(method = "loess",  se = FALSE) +
labs(x = "Fresh", y = "Grocery",
         title = "Relationship between spending of Fresh and Grocery goods",
         subtitle = "Very little relationship between spending of Fresh and Grocery products",
         caption = "Source:Customers")
```
## Interpretation
Region 1, Lisbon, between the 3, even if it is very subtle, show a negative relation between grocery and fresh products.
Region 2, Oporto, includes customers that spend more on fresh products than in grocery since the observations are more clustered to the right.
These 2 observations might be because these cities are near the ocean which allow people to have more fresh products.
Region 3 , has far more observations than the others and also they are more spread out which makes it harder to see a relationship. Besides that it has the highest amount of outliers, which can help us infer that the region include multiple cities and that is why the costumers' responses have more variety.

## Improvements 
Thanks to the feedback I received I scaled the y-axis too instead of just the x-axis and added the dollar sign in both. I also changed the subtitle so it can represent the primary insight of my chart. Finally, I changed some pieces of my interpretation to explain better the content and give more understandable insights. 

## End 
