---
title: "Lab 9"
author: "Gabriela CHajon"
format:
  html: 
    code-fold: false
    code-overflow: wrap
execute: 
  warning: false
  messages: false
  echo: true
toc: false
toc-location: left  
---

## Reflection

When it comes to data visualization, it is crucial to maintain honest results and transparency, from a Christian position. Regardless of the quality of the data, any misleading information displayed in graphs is immoral. Some might think that because the misinformation in Data Visualization can be in graphs and simpler things like that it should not be taken as a big concern. However, because of this same reason, graphs and images are easier to read and more appealing, so it can reach a greater audience which in the case of them being incorrect can lead to more false information. 

According to Camba et al. (2022), using homogenized data, avoiding from quoting data out of context, and making sure that graphs appropriately portray the data dimensions they represent are some of the fundamental ideas of maintaining graphical integrity.

From a Christian perspective, such actions go against the values of being humble, caring for others, and honesty. The biblical command to act justly and honestly in every aspect of life is in complete contrast to false information. The Bible places a strong emphasis on speaking the truth, which is something to keep in mind, particularly if providing false information could hurt other people and be a harm to 

Some of the sources of bias found in data visualization are usually companies or organizations intentions. Groups that are currently doing studies or reports with the intention of just making profits can have various motives to change the data. If the information is portrayed in a more favorable way for the organization they will get more benefits, in most cases, money. Other sources include specific beliefs of a group or part of society that is presenting the information. Instead of accurately representing the data in an objective manner, they might manipulate the presentation in order to promote their own beliefs.

As previously mentioned, there can be serious and far-reaching consequences if prejudice, inaccuracy, and dishonesty are allowed in data visualization. More importantly, dishonest data visualization techniques damage the audience's confidence and credibility, which strains the relationship between the communicator and the audience. Beyond the specific situation at hand, this integrity violation has wider ramifications since it undermines morality and ethics in general.
There are many examples of this in the bible, highlighting the importance of acting righteous and with integrity. Proverbs 11:says for example, "The integrity of the upright guides them, but the unfaithful are destroyed by their duplicity." 




## DATA

```{r}
library(ggplot2)
library(socviz)
p <- ggplot(gss_sm, aes(x=bigregion)) 
p + geom_bar() + 
	  labs(title = "Figure 4.6: A bar chart.") 
```

```{r}
library(ggplot2)
gss_sm <- data.frame(bigregion = c("Northeast", "Midwest", "South", "West"),
                     count = c(500, 520, 550, 570))
new_barchart <- ggplot(gss_sm, aes(x = bigregion, y = count)) +
                    geom_bar(stat = "identity", width = 0.5, fill = "skyblue") + 
                    geom_text(aes(label = ifelse(bigregion == "Northeast", "600", 
                                                 ifelse(bigregion == "Midwest", "590",
                                                        ifelse(bigregion == "South", "610",
                                                               ifelse(bigregion == "West", "620", ""))))), 
                              vjust = -0.5, size = 4) + 
                    labs(title = "Figure 4.6: A Misleading Bar Chart", 
                         x = "Regions", 
                         y = "Number of Participants (inaccurate scale)") + 
                    theme_minimal()

new_barchart

```

In this new bar chart, some changes were made to mislead viewers, there are based the data presented from the article "Examining data visualization pitfalls in scientific publications" by Nguyen et al(2021).
Firstly, because it was mentioned that misinformation resulting from size violations is one of the most frequent causes of false information, the bars' widths have been changed to seem closer together, making the discrepancies between the values of the areas appear less than they actually are. Additionally the article also remarks how easy it is to create misleading graphs when using incorrect labelling because it intentionally misleads viewers into drawing incorrect conclusions about the relative magnitudes of the regions' values. The changes made to the bar chart serve as an example of typical things, that might result in the data visualization process spreading false information. Beyond the immediate context, deceptive data visualization techniques can have long-term effects by fostering a culture of deceit and undermining the reliability of data especially for organizations, companies or even individuals analyzing data.




## References

Nguyen, V.T., Jung, K. & Gupta, V. Examining data visualization pitfalls in scientific publications. Vis. Comput. Ind. Biomed. Art 4, 27 (2021). https://doi.org/10.1186/s42492-021-00092-y

J. Camba, P. Company and V. Byrd, "Identifying Deception as a Critical Component of Visualization Literacy" in IEEE Computer Graphics and Applications, vol. 42, no. 01, pp. 116-122, 2022.
doi: 10.1109/MCG.2021.3132004





